ODGI: scalable tools for pangenome graphs

### 2021-06-02 manuscript/main.tex
edits

\newcommand{\FIXME}{\textbf{!!FIXME!!}}
\newcommand{\cmd}[1]{\textbf{\textrm{#1}}}
\newcommand{\cmdbf}[1]{{\textbf{#1}}}
\newcommand{\topic}[1]{{\cmdbf{#1}}:}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
      shortcomings of mainline genomics, in particular
      with regard to reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that implements new algorithms and has an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes in memory with fast
        parallel execution.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graph
        manipulation and visualisation. \odgi\ also provides an API that allows
        for writing new tool.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and documentation can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu
        % \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
        }
    \maketitle

@@

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are being produced at a high rate, including
    efforts from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}.  In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for homologs, paralogs and
    orthologs. So-called paths represent contigs that run through the
    nodes. A path can represent an individual chromosome or part
    thereof.  A bi-directed graph can even contain both strands of DNA
    as paths. A graph-based approach is particularly relevant for
    exploring complex variation and highly repetitive regions.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
    large data (see Table~\ref{tab:02}). We
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
    graph structure in computer memory (RAM) that can be dynamically updated using
    multiple CPU cores in parallel.

@@

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ includes the following tools:

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
    validate paths respect the graph topology.

@@

    Pangenome graphs stand to become a ubiquitous tool in
    genomics~\citep{Eizenga:2020}.  With \odgi\ we implemented a
    state-of-the-art tool suite, that can transform, analyse and
    visualise pangenome graphs at scale.  \odgi\ also exposes an API
    that can be used by other tools, e.g., for genotyping, pangenome
    graph building and normalization. Future work will add support for
    RNA and protein sequences and expand on metadata capabilities of
    large pangenome graphs.

@@

        %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[h!]
        \caption{ Example of performance improvement between
            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
            commands with output file sizes. The centromere region was
            extracted from CHR 6 of the HPRC year one assembly, i.e.,
            88~haploid, phased human genome assemblies from
            44~individuals plus the chm13 cell line and GRCh38
            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
            %
            First GFS was converted from GFA to \odgi\ and
            \vg\ formats, respectively.  \odgi's path implementation
            can process paths in parallel, outperforming \vg. For the
            same reason \cmd{odgi extract} is $20\times$ faster than
            \cmd{vg chunk} extracting a portion from the full
            graph. \FIXME: vg never finished.
            %
            All timings were performed on a server class machine
            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME}
        \begin{tabular}{{@{}lllll@{}}}
          \toprule
          Operation & Command & Runtime & Memory  & File size     \\
                    &      & (mm:ss) & (GB) & (GB)           \\
          \midrule
          Covert GFA to native format                         &       &       &                   \\
          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
          Extract subgraph                         &       &       &                   \\
          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
          \botrule
        \end{tabular}
        \label{tab:02}

>>>>

\newcommand{\FIXME}{\textbf{!!FIXME!!}}
\newcommand{\cmd}[1]{\textrm{#1}}
\newcommand{\cmdbf}[1]{{\textbf{#1}}}
\newcommand{\topic}[1]{{\cmdbf{#1}}:}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
      shortcomings of mainline genomics, in particular
      with regard to reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that implements new algorithms and has an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes in memory with fast
        parallel execution.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graph
        manipulation and visualisation. \odgi\ also provides an extensible API that allows
        for writing new tool.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and documentation can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
        }
    \maketitle

@@

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are being produced at a high rate, including
    efforts from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}.  In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for homologs, paralogs and
    orthologs. So-called paths represent reads, contigs, haplotypes, individual chromosomes or part thereof, that run through the
    nodes.   A bi-directed graph can even contain both strands of DNA
    as paths. A graph-based approach is particularly relevant for
    exploring complex variation and highly repetitive regions.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but does not scale for
    large data (see Table~\ref{tab:02}). We
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
    graph structure in computer memory that can be dynamically updated using
    multiple CPU cores in parallel.

@@

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ currently comes with over 30
    tools, including:

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
    it validates paths with respect the graph topology.

@@

    Pangenome graphs stand to become a ubiquitous tool in
    genomics~\citep{Eizenga:2020}.  With \odgi\ we implemented a
    state-of-the-art tool suite, that can transform, analyse and
    visualise pangenome graphs at scale.  \odgi\ also exposes an extensible API
    that can be used by other tools, e.g., for genotyping, pangenome
    graph building and normalization. Future work will add support for
    RNA and protein sequences and expand on metadata capabilities of
    large pangenome graphs.

@@

        %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[h!]
        \caption{ Example of performance improvement between
            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
            commands with output file sizes. The centromere region was
            extracted from CHR 6 of the HPRC year one assembly, i.e.,
            88~haploid, phased human genome assemblies from
            44~individuals plus the chm13 cell line and GRCh38
            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
            %
            First GFA was to converted to xg and next to \odgi\ and
            \vg\ formats respectively.  \odgi's path implementation
            can process paths in parallel, outperforming \vg. For the
            same reason \cmd{odgi extract} is $20\times$ faster than
            \cmd{vg chunk} extracting a portion from the full
            graph. \FIXME: vg never finished.
            %
            All timings were performed on a server class machine
            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME}
        \begin{tabular}{{@{}lllll@{}}}
          \toprule
          Operation & Command & Runtime & Memory  & File size     \\
                    &      & (mm:ss) & (GB) & (GB)           \\
          \midrule
          Covert GFA to native format                         &       &       &                   \\
          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
          Extract subgraph                         &       &       &                   \\
          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
          \botrule
        \end{tabular}
        \label{tab:02}
